Choosing Color Palettes for Statistical Graphics

### 2007-11-07 pkg/paper/Palettes/CSDA.Rnw
minor changes to CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@27 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages allow practicioners
to easily create color graphics which are therefore virtually omnipresent in electronic
publications such as technical reports, presentation slides, or the electronic
version of journal articles (e.g., in \emph{Computational Statistics \& Data Analysis})
and increasingly also in printed journals. However, more often than not,
color choice in such displays is sub-optimal because there is relatively
little guidance about how to choose appropriate colors for a particular visualization
task---and doing so is not a trivial task. There are three main obstacles to overcome:

\paragraph{The colors should not be unappealing.}
It is not necessary for the colors in a statistical plot to reflect the latest
in fashion trends, but basic principles such as avoiding fully saturated colors
should be adhered to. The requirement is not that the user should have
a degree in graphic design, but that the software should provide users with an
intuitive way to select colors and control their basic properties such as how
bright the colors are. Thus, it is necessary to employ a \emph{color model}
or \emph{color space} that describes colors in terms of their perceptual
properties hue, brightness, and colorfulness.

A color model typically supported by software packages is the specification
of colors as Red-Green-Blue (RGB) triplets. However, this specification
does not correspond to human color perception but to color generation on
a computer screen where different color sensations are created by
by mixing different amounts of the primary colors red, green and blue
\citep[see][for more details]{color:Poynton:2000}. For humans, it is
virtually impossible to control the perceptual properties of a color in this
color space because there is no single dimension that corresponds to, for example,
the hue or the brightness of the color. Therefore, various perceptually based
color spaces have been suggested where each dimension of the color space can
be matched with a perceptual property. One approach widely implemented in
software packages are Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
a simple transformation of RGB triplets \citep[see]{color:Wiki+HSV:2007}.
Unfortunately, the dimensions in HSV space map poorly to perceptual
properties---furthermore, the use of HSV colors encourages the use of flashy and
highly saturated colors which are good for drawing attention to a plot but hard
to look at for a longer time. A perceptually based color model that avoids
these problems are Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
Albeit being less commonly available in software packages, it is extremely
useful for sepecifying colors intuitively.

@@

A natural solution to this task is to vary (at least) one perceptual property
of the colors, e.g., the hue or the brightness, keeping other properties
fixed. In a perceptually based color space, this corresponds to selecting
colors by traversing paths through this space along its dimensions.
This approach is implemented in many color picker tools
\citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
however, most commonly based on the HSV model.
As argued above, the dimensions in HSV space are not truly independent and
hence it is not possible to vary just one perceptual property while keeping
the others fixed. For example, a change in hue results in a color that may also be
perceptually brighter and more saturated. This means that it is relatively
difficult to select sets of HSV coordinates that yield colors that are
``in harmony'' \citep[see][]{color:Munsell:1905}. For statistical graphics,
this is important because it can introduce size distortions in the perception
of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
These problems can again be addressed by employing HCL colors as a more
appropriate perceptually based color model. Above, we argued that HCL space
is useful because it allows humans to understand where within the space
a particular HCL triplet is located. Here, we extend this argument: HCL also allows
us to understand where we are going, i.e., motion within the color space and
distances between colors  have an intuitive and rational meaning.

@@

An approach exemplary in this respect is \pkg{ColorBrewer.org}
\citep{color:Harrower+Brewer:2003}, an online tool for selecting color
schemes for maps. It provides a rich collection of prefabricated palettes
(with a fixed maximal number of colors), supplies information about the
suitability for color-blind viewers or grayscale printing, and guides its
users how to choose a suitable palette for coding various types of
information. The drawback to this tool is that it only provides a fixed
set of colors for each palette and there is no way to extend the
existing palettes.

>>>>

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. 
It is easy to create color graphics with any statistical software 
pacakge and color images are therefore virtually omnipresent in electronic
publications such as technical reports, presentation slides, or the electronic
version of journal articles (e.g., in \emph{Computational Statistics \& Data Analysis})
and increasingly also in printed journals. However, more often than not,
color choice in such displays is sub-optimal because there is relatively
little guidance about how to choose appropriate colors for a particular visualization
task---and doing so is not a trivial task. 
When selecting colors for a statistical graphic,
there are three main obstacles to overcome:

\paragraph{The colors should not be unappealing.}
It is not necessary for the colors in a statistical plot to reflect the latest
in fashion trends, but basic principles such as avoiding fully saturated colors
should be adhered to. The requirement is not that the user should have
a degree in graphic design, but that the software should provide users with an
intuitive way to select colors and control their basic properties, such as how
bright the colors are. Thus, it is necessary to employ a \emph{color model}
or \emph{color space} that describes colors in terms of their perceptual
properties: hue, brightness, and colorfulness.

A color model typically supported by software packages is the specification
of colors as Red-Green-Blue (RGB) triplets. However, this specification
does not correspond to human color perception but to color generation on
a computer screen where different color sensations are created
by mixing different amounts of the primary colors red, green and blue
\citep[see][for more details]{color:Poynton:2000}. For humans, it is
virtually impossible to control the perceptual properties of a color in this
color space because there is no single dimension that corresponds to, for example,
the hue or the brightness of the color. Therefore, various perceptually based
color spaces have been suggested where each dimension of the color space can
be matched with a perceptual property. One approach widely implemented in
software packages are Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
a simple transformation of RGB triplets \citep[see]{color:Wiki+HSV:2007}.
Unfortunately, the dimensions in HSV space map poorly to perceptual
properties---furthermore, the use of HSV colors encourages the use of flashy and
highly saturated colors which are good for drawing attention to a plot but hard
to look at for a longer time. A perceptually based color model that avoids
these problems involves
Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
which is
a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
Although this is less commonly available in software packages, it is extremely
useful for sepecifying colors intuitively.

@@

A natural solution for this task is to vary (at least) one perceptual property
of the colors, e.g., the hue or the brightness, keeping other properties
fixed. In a perceptually based color space, this corresponds to selecting
colors by traversing paths through this space along its dimensions.
This approach is implemented in many color picker tools
\citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
however, most are commonly based on the HSV model.
As argued above, the dimensions in HSV space are not truly independent and
hence it is not possible to vary just one perceptual property while keeping
the others fixed. For example, a change in hue results in a color that may also be
perceptually brighter and more saturated. This means that it is relatively
difficult to select sets of HSV coordinates that yield colors that are
``in harmony'' \citep[see][]{color:Munsell:1905}. For statistical graphics,
this is important because it can introduce size distortions in the perception
of shaded areas and can produce
 optical illusions \citep{color:Cleveland+McGill:1983}.
These problems can again be addressed by employing HCL colors as a more
appropriate perceptually based color model. Above, we argued that HCL space
is useful because it allows humans to understand where within the space
a particular HCL triplet is located. Here, we extend this argument: HCL also allows
us to understand where we are going, i.e., motion within the color space and
distances between colors  have an intuitive and rational meaning.

@@

An example of this approach is \pkg{ColorBrewer.org}
\citep{color:Harrower+Brewer:2003}, an online tool for selecting color
schemes for maps. It provides a rich collection of prefabricated palettes
(with a fixed maximal number of colors), supplies information about the
suitability for color-blind viewers or grayscale printing, and guides its
users how to choose a suitable palette for coding various types of
information. The drawback to this tool is that it only provides a fixed
set of colors for each palette and there is no way to extend the
existing palettes.
